Deciding Bitvector Termination with Program Synthesis

### 2014-07-04 papers/termination/paper.tex
Deterministic completeness

\begin{figure*}
\begin{framed}
\begin{definition}[Second-order Unconditional Termination Formula {\bf [UT]}]
\label{def:termination-formula}
\begin{align*}
 \exists R . \forall x, x' . & G(x) \wedge T(x, x') \rightarrow R(x) > 0 \wedge R(x) > R(x')
\end{align*}
\end{definition}

\begin{definition}[Non-Termination Formula - Open Recurrence Set  {\bf [ONT]}]
\label{def:nonterm-formula}
 \begin{align*}
  \exists N, x_0 . \forall x . \exists x' . & N(x_0) ~ \wedge ~ N(x) \rightarrow G(x) ~ \wedge \\
							& N(x) \rightarrow T(x, x') \wedge N(x') 
 \end{align*}
\end{definition}

\begin{definition}[Non-Termination Formula - Closed Recurrence Set {\bf [CNT]}]
\label{def:deterministic-nonterm-formula}
 \begin{align*}
  \exists N, x_0 . \forall x, x' . & N(x_0) ~ \wedge ~ N(x) \rightarrow G(x) ~ \wedge \\
							& N(x) \wedge T(x, x') \rightarrow N(x') 
 \end{align*}
\end{definition}

@@

\noindent {\bf Unconditional termination.}
To show that $L$ is unconditionally terminating (i.e. it eventually terminates regardless of the
state it starts in), we can find a ranking function for $L$.
The existence of a ranking function is equivalent to the satisfiability
of the formula from Definition~\ref{def:termination-formula}, so a witness to the satisfiability
of formula {\bf [UT]} is also a ranking function and thus a proof of $L$'s unconditional termination.

@@

\begin{definition}[Conditional Termination Formula {\bf [CT]}]
\label{def:conditional-termination-formula}
 \begin{align*}
  \exists R, W . \forall x, x' . & I(x) \wedge G(x) \rightarrow W(x) ~ \wedge \\
                                 & G(x) \wedge W(x) \wedge T(x, x') \rightarrow W(x') \wedge R(x) > 0 
  \wedge R(x) > R(x')
 \end{align*}
\end{definition}

@@

Conditional termination is then equivalent to the
formula {\bf [CT]} of Defintion~\ref{def:conditional-termination-formula}. If this formula is satisfiable,
two witnesses are returned:
\begin{itemize}
\item $W$ is an inductive invariant of $L$ that is established by the initial states $I$ if the loop
guard $G$ is met.
\item $R$ is a ranking function for $L$ as restricted by $W$ -- that is to say, $R$ need only
be well founded on those states satisfying $W \wedge G$.  Since $W$ is an inductive invariant of $L$,
$R$ is strong enough to show that $L$ terminates from any of its initial states.
\end{itemize}

@@

\section{Soundness, Completeness and Complexity}
\begin{theorem}
 {\sc Headshot} is sound -- if it says a program terminates then it does.
\end{theorem}

\begin{proof}
 The ranking function constitutes a constructive, machine checked proof that the program terminates.
\end{proof}

\begin{theorem}
 {\sc Headshot} is complete for terminating programs -- if a program terminates, {\sc Headshot} will find a proof that it does.
\end{theorem}

\begin{proof}
 If the program under analysis $P$ has variables $v_1, \ldots, v_k$ each of which is $b$ bits wide, its state space $\mathcal{S}$ is of size $2^{bk}$.
 A ranking function $R: \mathcal{S} \to \mathcal{D}$ for $P$ exists iff $P$ terminates.  Since $R$ is injective, we have that
 $\| \mathcal{D} \| \geq \| \mathcal{S} \|$.  If $\| \mathcal{D} \| > \| \mathcal{S} \|$, we can construct a function $R': \mathcal{S} \to \mathcal{D'}$
 with $ \| \mathcal{D'} = \| \mathcal{S} \|$ by just setting $R' = R|_\mathcal{S}$, i.e. $R'$ is just the restriction of $R$ to $\mathcal{S}$
 Since $\mathcal{S}$ already comes equipped with a natural well ordering we can also construct $R'' = \iota \circ R'$
 where $\iota: \mathcal{D'} \to \mathcal{S}$ is the unique order isomorphism from $\mathcal{D'}$ to $\mathcal{S}$.
 So assuming that $P$ terminates, there is some ranking function $R''$ that is just a permutation of $\mathcal{S}$.

 It is fairly easy to see that any finite permutation is computed by a finite program in the proof language generated
 by {\sc Kalashnikov}.  A very inefficient construction which computes the permutation $R(x) = x+1$ for one variable
 is:

 \begin{verbatim}
t1 = 0
t2 = v1 == 0
t3 = ITE(t2, 1, t1)
t4 = v1 == 1
t5 = ITE(t4, 2, t3)
...
 \end{verbatim}

If the number of variables $k > 1$ then in general the ranking function will be lexicographic with dimension $\leq k$.

@@

\begin{theorem}
 {\sc Headshot} is a decision procedure -- it always terminates.
\end{theorem}

>>>>

\begin{figure*}
\begin{framed}
\begin{definition}[Second-order Unconditional Termination Formula {\bf [UT]}]
\label{def:UT}
\begin{align*}
 \exists R . \forall x, x' . & G(x) \wedge T(x, x') \rightarrow R(x) > 0 \wedge R(x) > R(x')
\end{align*}
\end{definition}

\begin{definition}[Non-Termination Formula - Open Recurrence Set  {\bf [ONT]}]
\label{def:ont}
 \begin{align*}
  \exists N, x_0 . \forall x . \exists x' . & N(x_0) ~ \wedge ~ N(x) \rightarrow G(x) ~ \wedge \\
							& N(x) \rightarrow T(x, x') \wedge N(x') 
 \end{align*}
\end{definition}

\begin{definition}[Non-Termination Formula - Closed Recurrence Set {\bf [CNT]}]
\label{def:cnt}
 \begin{align*}
  \exists N, x_0 . \forall x, x' . & N(x_0) ~ \wedge ~ N(x) \rightarrow G(x) ~ \wedge \\
							& N(x) \wedge T(x, x') \rightarrow N(x') 
 \end{align*}
\end{definition}

@@

\noindent {\bf Unconditional termination.}
To show that $L$ is unconditionally terminating (i.e. it eventually terminates regardless of the
state it starts in), we can find a ranking function for $L$.
The existence of a ranking function is equivalent to the satisfiability
of the formula from Definition~\ref{def:UT}, so a witness to the satisfiability
of formula {\bf [UT]} is also a ranking function and thus a proof of $L$'s unconditional termination.

@@

\begin{definition}[Conditional Termination Formula {\bf [CT]}]
\label{def:ct}
 \begin{align*}
  \exists R, W . \forall x, x' . & I(x) \wedge G(x) \rightarrow W(x) ~ \wedge \\
                                 & G(x) \wedge W(x) \wedge T(x, x') \rightarrow W(x') \wedge R(x) > 0 
  \wedge R(x) > R(x')
 \end{align*}
\end{definition}

@@

Conditional termination is then equivalent to the
formula {\bf [CT]} of Defintion~\ref{def:ct}. If this formula is satisfiable,
two witnesses are returned:
\begin{itemize}
\item $W$ is an inductive invariant of $L$ that is established by the initial states $I$ if the loop
guard $G$ is met.
\item $R$ is a ranking function for $L$ as restricted by $W$ -- that is to say, $R$ need only
be well founded on those states satisfying $W \wedge G$.  Since $W$ is an inductive invariant of $L$,
$R$ is strong enough to show that $L$ terminates from any of its initial states.
\end{itemize}

@@

\section{Soundness, Completeness and Complexity}
\begin{theorem}
\label{thm:sound}
 {\sc Headshot} is sound -- if it returns a verdict of termination, the program under
 analysis terminates on all inputs.  Similarly if it returns a verdict of non-termination,
 the program under analysis has at least one infinite execution.
\end{theorem}

\begin{proof}
 {\sc Headshot}'s output constitutes a constructive, machine checked proof of (non-)termination.
\end{proof}

\begin{theorem}
\label{thm:complete-termination}
 {\sc Headshot} is complete for terminating programs -- if a program terminates, {\sc Headshot} will find a proof that it does.
\end{theorem}

\begin{proof}
 If the program under analysis $P$ has variables $v_1, \ldots, v_k$ each of which is $b$ bits wide, its state space $\mathcal{S}$ is of size $2^{bk}$.
 A ranking function $R: \mathcal{S} \to \mathcal{D}$ for $P$ exists iff $P$ terminates.  Since $R$ is injective, we have that
 $\| \mathcal{D} \| \geq \| \mathcal{S} \|$.  If $\| \mathcal{D} \| > \| \mathcal{S} \|$, we can construct a function $R': \mathcal{S} \to \mathcal{D'}$
 with $ \| \mathcal{D'} = \| \mathcal{S} \|$ by just setting $R' = R|_\mathcal{S}$, i.e. $R'$ is just the restriction of $R$ to $\mathcal{S}$
 Since $\mathcal{S}$ already comes equipped with a natural well ordering we can also construct $R'' = \iota \circ R'$
 where $\iota: \mathcal{D'} \to \mathcal{S}$ is the unique order isomorphism from $\mathcal{D'}$ to $\mathcal{S}$.
 So assuming that $P$ terminates, there is some ranking function $R''$ that is just a permutation of $\mathcal{S}$.
 If the number of variables $k > 1$ then in general the ranking function will be lexicographic with dimension $\leq k$
 and each co-ordinate of the output being a single $b$-bit value.

 It is fairly easy to see that any finite permutation is computed by a finite program in the proof language generated
 by {\sc Kalashnikov}.  A very inefficient construction which computes the permutation $R(x) = x+1$ for one variable
 is:
 \begin{verbatim}
t1 = 0
t2 = v1 == 0
t3 = ITE(t2, 1, t1)
t4 = v1 == 1
t5 = ITE(t4, 2, t3)
...
 \end{verbatim}
It is easy to see how this generalises to computing an arbitrary permutation.

@@

\begin{theorem}
 {\sc Headshot} is complete for deterministic programs -- {\sc Headshot} will terminate when run on
 any program with a deterministic transition relation.
\end{theorem}

\begin{proof}
 If the program under analysis is terminating, then by Theorem~\ref{thm:complete-termination} {\sc Headshot}
 will terminate having computed a ranking function.

 Conversely if the program is non-terminating, {\sc Headshot} will terminate
 having computed a recurrence set.  Since, by assumption, the program under analysis is determinstic,
 Definition~\ref{def:ont} is equivalent to Definition~\ref{def:cnt}.
 That is to say, each open recurrence set is also a closed recurrence set.  Since the specification for a predicate
 to encode a closed recurrence set is in the synthesis fragment, {\sc Kalashnikov} will find a program computing
 such a predicate if one exists.  The proof is similar to the proof of Theorem~\ref{thm:complete-termination}.
\end{proof}
